---
title: '508 Final: Heroin Overdose Prediction'
author: "Jingyi Li & Sihan Ren"
date: "2023-11-29"
output: html_document
---

## Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(RSocrata)
library(viridis)
library(spatstat.explore)
library(raster)
library(spdep)
library(FNN)
library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)
library(tidycensus)
library(classInt)

```

## Project

|    Predict drug overdose events to better allocate prevention resources in Mesa.

## Use case

|    Use the predictions generated by the model to identify areas or demographics with a higher risk of drug overdose. Raising more attention of the healthcare professionals, public health agencies, law enforcement, and community organizations on these high-risk areas. Also urging Policy makers to allocate prevention resources, such as outreach programs, naloxone distribution, and treatment facilities, to these high-risk areas.

## Methods

(1) Find Data: crime, grafitti, homelessness, income, unemployment, race, education level as predictors od drug overdose.

(2) Run models(OLS?GWR?Spatial?) to explore the relationship between the predictors and dependent vairable

(3) Validate model accuracy

(4) Analyze the result and conclude with feasibility of the model

## Data

```{r cars}
drugOverdose<-read.csv("DATA/Fire_and_Medical_Opioid_Overdose_Incidents.csv")%>%
  mutate(X = as.numeric(Longitude),Y = as.numeric(Latitude)) %>% 
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant")%>%
    st_transform('EPSG:2223')

mesaBoundary<- st_read("https://services2.arcgis.com/1gVyYKfYgW5Nxb1V/arcgis/rest/services/CityBoundary/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")%>%
  st_transform('EPSG:2223')

census20<-  get_acs(geography = "tract",
          variables = c("B19013_001E",# medHH income
                        "B01003_001E",#total pop
                        "B15003_022E",# bachelor or more
                        "B23025_005E",# unemployed
                        "B02001_002E",# white alone
                        "B02001_003E"),#black alone
          year=2020, state=04,
          county=13, geometry=TRUE) %>% 
  st_transform('EPSG:2223')

census20<-census20%>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B01003_001, 
         MedHHInc = B19013_001, 
         BachMore = B15003_022,
         Unemploy = B23025_005,
         BlackAlone = B02001_003,
         WhiteAlone = B02001_002
         )

census20<- 
  census20 %>%
  mutate(pctUnemploy = ifelse(TotalPop > 0, Unemploy / TotalPop, 0),
         pctWhite = ifelse(TotalPop > 0, (WhiteAlone / TotalPop), 0),
         pctBlack = ifelse(TotalPop > 0, (BlackAlone / TotalPop), 0),
         pctBachMore= ifelse(TotalPop>0, (BachMore / TotalPop), 0),
         year = "2020") %>%
  dplyr::select(-BachMore,-Unemploy,-BlackAlone,-WhiteAlone)
```

## Exploratory analysis

### Point and density graph of drug overdose

```{r echo=TRUE,results='hide', message=FALSE, warning=FALSE}
filtered_Overdose <- st_intersection(drugOverdose, mesaBoundary)%>%filter(Year=="2020")
grid.arrange(ncol=2,
ggplot() + 
  geom_sf(data = mesaBoundary) +
  geom_sf(data = filtered_Overdose, colour="red", size=0.1, show.legend = "point") +
  labs(title= "Drug Overdose, Mesa - 2020") +
  theme(plot.title = element_text(size = 14),axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank()),

ggplot() + 
  geom_sf(data = mesaBoundary, fill = "grey40") +
  stat_density2d(data = data.frame(st_coordinates(filtered_Overdose)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_viridis() +
  scale_alpha(range = c(0.00, 0.35), guide = FALSE) +
  labs(title = "Density of Drug Overdose") +
  theme(plot.title = element_text(size = 14),axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank()) + theme(legend.position = "none"))
```

### Creating Fishnet 

```{r fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
fishnet <- 
  st_make_grid(mesaBoundary,
               cellsize = 1500, 
               square = TRUE) %>%
  .[mesaBoundary] %>%            # fast way to select intersecting polygons
  st_sf() %>%
  mutate(uniqueID = 1:n())
fishnet
```

### Join Points to Fishnet  

```{r}
drugNet <- 
  dplyr::select(filtered_Overdose) %>% 
  mutate(countOverdose = 1) %>% 
  aggregate(., fishnet, sum) %>%
  mutate(countOverdose = replace_na(countOverdose, 0),
         uniqueID = rownames(.),
         cvID = sample(round(nrow(fishnet) / 24), size=nrow(fishnet), replace = TRUE))

ggplot() +
  geom_sf(data = drugNet, aes(fill = countOverdose)) +
  scale_fill_viridis() +
  labs(title = "Count of Drug Overdose for the Fishnet")+
  theme_minimal()
```
### Drug Overdose count histogram

```{r}
ggplot(drugNet, aes(x = countOverdose)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Overdose Histogram",
       x = "Overdose Count",
       y = "Frequency")
```

### Join Risk Factors into Fishnet

```{r}
#Risk Factors Data
crime<-read.socrata("https://data.mesaaz.gov/Police/Police-Incidents/39rt-2rfj")%>%
  mutate(year = substr(report_date,1,4)) %>% filter(year == "2020") %>%
    dplyr::select(Y = latitude, X = longitude) %>%
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
    st_transform(st_crs(fishnet)) %>%
    mutate(Legend = "crime")
graffiti<- read.csv("DATA/Transportation_Graffiti_20231129.csv")%>%
  mutate(year = substr(Date.Reported,1,4))%>%
    dplyr::select(Y = Latitude, X = Longitude) %>%
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
    st_transform(st_crs(fishnet)) %>%
    mutate(Legend = "graffiti")
homeless<-read.csv("DATA/Unsheltered_Point_in_Time__PIT__Count_Phoenix_Metro_Area_20231129.csv")%>%
  mutate(year = substr(Reporting.Date,1,4))%>%
    dplyr::select(Y = Latitude, X = Longitude) %>%
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
    st_transform(st_crs(fishnet)) %>%
    mutate(Legend = "homeless")
hospital<-st_read("DATA/Hospitals_031816.geojson")%>%
  filter(CITY=="MESA")%>%
  dplyr::select(geometry=geometry)%>%
  st_transform(st_crs(fishnet))%>%
  mutate(Legend="hospital")
park<-st_read("DATA/Park and Recreation Areas.geojson")%>%
  dplyr::select(geometry=geometry)%>%
  st_transform(st_crs(fishnet))%>%
  mutate(Legend="park")

```


```{r}
vars_net <- 
  rbind(crime,graffiti, homeless,hospital,park) %>%
  st_join(fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID, Legend) %>%
  summarize(count = n()) %>%
  full_join(fishnet) %>%
  spread(Legend, count, fill = 0) %>%
  st_as_sf() %>%
  dplyr::select(-`<NA>`) %>%
  na.omit() %>%
  ungroup()
```

### Plot fishnet of risk factors

```{r}
vars_net.long <- 
  gather(vars_net, Variable, value, -geometry, -uniqueID)

vars <- unique(vars_net.long$Variable)
mapList <- list()

for(i in vars){
  mapList[[i]] <- 
    ggplot() +
      geom_sf(data = filter(vars_net.long, Variable == i), aes(fill=value), colour=NA) +
      scale_fill_viridis(name="") +
      labs(title=i) +
      theme(axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank())}

do.call(grid.arrange,c(mapList, ncol=3, top="Risk Factors by Fishnet"))
```

### Nearest Neighbor of Risk Factors

```{r}
st_c    <- st_coordinates
st_coid <- st_centroid

crime_coordinates <- st_c(crime)
graffiti_coordinates <- st_c(graffiti)
homeless_coordinates <- st_c(homeless)
hospital_coordinates<-st_c(hospital)
park_coordinates<-st_c(park)
vars_net_coordinates <- st_c(st_coid(vars_net))
# create NN for risk factors
vars_net <-
  vars_net %>%
    mutate(
      crime.nn =
        nn_function(vars_net_coordinates, crime_coordinates,3),
      graffiti.nn =
        nn_function(vars_net_coordinates, graffiti_coordinates,3),
      homeless.nn =
        nn_function(vars_net_coordinates, homeless_coordinates,3),
      hospital.nn =
        nn_function(vars_net_coordinates, hospital_coordinates,3),
      park.nn =
        nn_function(vars_net_coordinates, park_coordinates,3)
      )

#plot multiple maps using for loop
vars_net.long.nn <- 
  dplyr::select(vars_net, ends_with(".nn")) %>%
    gather(Variable, value, -geometry)

vars <- unique(vars_net.long.nn$Variable)
mapList <- list()

for(i in vars){
  mapList[[i]] <- 
    ggplot() +
      geom_sf(data = filter(vars_net.long.nn, Variable == i), aes(fill=value), colour=NA) +
      scale_fill_viridis(name="",trans = "reverse") +
      labs(title=i) +
      theme(plot.title = element_text(size = 14),axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank())}

do.call(grid.arrange,c(mapList, ncol = 3, top = "Nearest Neighbor risk Factors by Fishnet"))
```
### Local Moran's I

```{r}
tract<-read_sf("DATA/Mesa Census Tracts To City Boundary/tracts.shp")%>%
  st_transform(st_crs(fishnet))

#create final net
vars_net$uniqueID <- as.character(vars_net$uniqueID)
drugNet$uniqueID <- as.character(drugNet$uniqueID)
final_net <-
  left_join(drugNet, st_drop_geometry(vars_net), by="uniqueID") 

#spacially join final net to tracts
final_net <-
  st_centroid(final_net) %>%
    st_join(dplyr::select(tract, tractce)) %>%
      st_drop_geometry() %>%
      left_join(dplyr::select(final_net, geometry, uniqueID)) %>%
      st_sf() %>%
  na.omit()

## make polygon to tracts 
final_net.nb <- poly2nb(as_Spatial(final_net), queen=TRUE)
## make tracts to list of weigths
final_net.weights <- nb2listw(final_net.nb, style="W", zero.policy=TRUE)

local_morans <- localmoran(final_net$countOverdose, final_net.weights, zero.policy=TRUE) %>% as.data.frame()
# join local Moran's I results to fishnet
final_net.localMorans <- 
  cbind(local_morans, as.data.frame(final_net)) %>% 
  st_sf() %>%
  dplyr::select(Overdose_Count = countOverdose, 
                Local_Morans_I = Ii, 
                P_Value = `Pr(z != E(Ii))`) %>%
  mutate(Significant_Hotspots = ifelse(P_Value <= 0.05, 1, 0)) %>%
  gather(Variable, Value, -geometry)

#plot  
vars <- unique(final_net.localMorans$Variable)
varList <- list()

for(i in vars){
  varList[[i]] <- 
    ggplot() +
      geom_sf(data = filter(final_net.localMorans, Variable == i), 
              aes(fill = Value), colour=NA) +
      scale_fill_viridis(name="") +
      labs(title=i) +
      mapTheme() + theme(legend.position="bottom")
}

do.call(grid.arrange,c(varList, ncol = 4, top = "Local Morans I statistics, Drug Overdose"))
```


### Correlation

```{r fig.height=12,fig.width=8}
vars_net$uniqueID <- as.numeric(vars_net$uniqueID)
drugNet$uniqueID <- as.numeric(drugNet$uniqueID)
final_net <-
  left_join(drugNet, st_drop_geometry(vars_net), by="uniqueID") 
correlation.long <-
  st_drop_geometry(final_net) %>%
    dplyr::select(-uniqueID, -cvID) %>%
    gather(Variable, Value, -countOverdose)

correlation.cor <-
  correlation.long %>%
    group_by(Variable) %>%
    summarize(correlation = cor(Value, countOverdose, use = "complete.obs"))
    
ggplot(correlation.long, aes(Value, countOverdose)) +
  geom_point(size = 0.1) +
  geom_text(data = correlation.cor, aes(label = paste("r =", round(correlation, 2))),
            x=-Inf, y=Inf, vjust = 1.5, hjust = -.1) +
  geom_smooth(method = "lm", se = FALSE, colour = "orange") +
  facet_wrap(~Variable, ncol = 2, scales = "free") +
  labs(title = "Overdose count as a function of risk factors") +
  plotTheme()
ggsave("correlation.jpg",height = 12, width=8)
```

