---
title: '508 Final: Heroin Overdose Prediction'
author: "Jingyi Li & Sihan Ren"
date: "2023-11-29"
output: html_document
---

## Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(RSocrata)
library(viridis)
library(spatstat.explore)
library(raster)
library(spdep)
library(FNN)
library(grid)
library(gridExtra)
library(knitr)
library(kableExtra)
library(tidycensus)
library(classInt)

options(tigris_class = "sf")
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
```

## Project

|    Predict drug overdose events to better allocate prevention resources in Mesa.

## Use case

|    Use the predictions generated by the model to identify areas or demographics with a higher risk of drug overdose. Raising more attention of the healthcare professionals, public health agencies, law enforcement, and community organizations on these high-risk areas. Also urging Policy makers to allocate prevention resources, such as outreach programs, naloxone distribution, and treatment facilities, to these high-risk areas.

## Methods

(1) Find Data: crime, grafitti, homelessness, income, unemployment, race, education level as predictors od drug overdose.

(2) Run models(OLS?GWR?Spatial?) to explore the relationship between the predictors and dependent vairable

(3) Validate model accuracy

(4) Analyze the result and conclude with feasibility of the model

## Data

```{r cars}
drugOverdose<-read.csv("DATA/Fire_and_Medical_Opioid_Overdose_Incidents.csv")%>%
  mutate(X = as.numeric(Longitude),Y = as.numeric(Latitude)) %>% 
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant")%>%
    st_transform('EPSG:2223')

mesaBoundary<- st_read("https://services2.arcgis.com/1gVyYKfYgW5Nxb1V/arcgis/rest/services/CityBoundary/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson")%>%
  st_transform('EPSG:2223')

census20<-  get_acs(geography = "block group",
          variables = c("B19013_001E",# medHH income
                        "B01003_001E",#total pop
                        "B15003_022E",# bachelor or more
                        "B23025_005E",# unemployed
                        "B02001_002E",# white alone
                        "B02001_003E"),#black alone
          year=2020, state=04,
          county=13, geometry=TRUE) %>% 
  st_transform('EPSG:2223')

census20<-census20%>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B01003_001, 
         MedHHInc = B19013_001, 
         BachMore = B15003_022,
         Unemploy = B23025_005,
         BlackAlone = B02001_003,
         WhiteAlone = B02001_002
         )

census20<- 
  census20 %>%
  mutate(pctUnemploy = ifelse(TotalPop > 0, Unemploy / TotalPop, 0),
         pctWhite = ifelse(TotalPop > 0, (WhiteAlone / TotalPop), 0),
         pctBlack = ifelse(TotalPop > 0, (BlackAlone / TotalPop), 0),
         pctBachMore= ifelse(TotalPop>0, (BachMore / TotalPop), 0),
         year = "2020") %>%
  dplyr::select(-BachMore,-Unemploy,-BlackAlone,-WhiteAlone)
```

## Exploratory analysis

### Point and density graph of drug overdose

```{r echo=TRUE,results='hide', message=FALSE, warning=FALSE}
filtered_Overdose <- st_intersection(drugOverdose, mesaBoundary)
grid.arrange(ncol=2,
ggplot() + 
  geom_sf(data = mesaBoundary) +
  geom_sf(data = filtered_Overdose, colour="red", size=0.1, show.legend = "point") +
  labs(title= "Drug Overdose, Mesa - 2020") +
  theme(plot.title = element_text(size = 14),axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank()),

ggplot() + 
  geom_sf(data = mesaBoundary, fill = "grey40") +
  stat_density2d(data = data.frame(st_coordinates(filtered_Overdose)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_viridis() +
  scale_alpha(range = c(0.00, 0.35), guide = FALSE) +
  labs(title = "Density of Drug Overdose") +
  theme(plot.title = element_text(size = 14),axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank()) + theme(legend.position = "none"))
```

### Creating Fishnet 

```{r fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
fishnet <- 
  st_make_grid(mesaBoundary,
               cellsize = 1500, 
               square = TRUE) %>%
  .[mesaBoundary] %>%            # fast way to select intersecting polygons
  st_sf() %>%
  mutate(uniqueID = 1:n())
fishnet
```

### Join Points to Fishnet  

```{r}
drugNet <- 
  dplyr::select(filtered_Overdose) %>% 
  mutate(countOverdose = 1) %>% 
  aggregate(., fishnet, sum) %>%
  mutate(countOverdose = replace_na(countOverdose, 0),
         uniqueID = rownames(.),
         cvID = sample(round(nrow(fishnet) / 24), size=nrow(fishnet), replace = TRUE))

ggplot() +
  geom_sf(data = drugNet, aes(fill = countOverdose)) +
  scale_fill_viridis() +
  labs(title = "Count of Drug Overdose for the Fishnet")+
  theme_minimal()
```

### Join Risk Factors into Fishnet

 
```{r}
#Risk Factors Data
crime<-read.socrata("https://data.mesaaz.gov/Police/Police-Incidents/39rt-2rfj")%>%
  mutate(year = substr(report_date,1,4)) %>% filter(year == "2020") %>%
    dplyr::select(Y = latitude, X = longitude) %>%
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
    st_transform(st_crs(fishnet)) %>%
    mutate(Legend = "crime")
graffiti<- read.csv("DATA/Transportation_Graffiti_20231129.csv")%>%
  mutate(year = substr(Date.Reported,1,4))%>%
    dplyr::select(Y = Latitude, X = Longitude) %>%
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
    st_transform(st_crs(fishnet)) %>%
    mutate(Legend = "graffiti")
homeless<-read.csv("DATA/Unsheltered_Point_in_Time__PIT__Count_Phoenix_Metro_Area_20231129.csv")%>%
  mutate(year = substr(Reporting.Date,1,4))%>%
    dplyr::select(Y = Latitude, X = Longitude) %>%
    na.omit() %>%
    st_as_sf(coords = c("X", "Y"), crs = 4326, agr = "constant") %>%
    st_transform(st_crs(fishnet)) %>%
    mutate(Legend = "homeless")
  

```


```{r}
vars_net <- 
  rbind(crime,graffiti, homeless) %>%
  st_join(fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID, Legend) %>%
  summarize(count = n()) %>%
  full_join(fishnet) %>%
  spread(Legend, count, fill = 0) %>%
  st_as_sf() %>%
  dplyr::select(-`<NA>`) %>%
  na.omit() %>%
  ungroup()
```
### Plot fishnet of risk factors

```{r}
vars_net.long <- 
  gather(vars_net, Variable, value, -geometry, -uniqueID)

vars <- unique(vars_net.long$Variable)
mapList <- list()

for(i in vars){
  mapList[[i]] <- 
    ggplot() +
      geom_sf(data = filter(vars_net.long, Variable == i), aes(fill=value), colour=NA) +
      scale_fill_viridis(name="") +
      labs(title=i) +
      theme(axis.title = element_blank(), axis.text = element_blank(),axis.ticks = element_blank())}

do.call(grid.arrange,c(mapList, ncol=3, top="Risk Factors by Fishnet"))
```

### Nearest Neighbor of Risk Factors

```{r}
st_c    <- st_coordinates
st_coid <- st_centroid

crime_coordinates <- st_c(crime)
graffiti_coordinates <- st_c(graffiti)
homeless_coordinates <- st_c(homeless)
vars_net_coordinates <- st_c(st_coid(vars_net))
# create NN for alley light out and bars
vars_net <-
  vars_net %>%
    mutate(
      crime.nn =
        nn_function(vars_net_coordinates, crime_coordinates,3),
      graffiti.nn =
        nn_function(vars_net_coordinates, graffiti_coordinates,3),
      homeless.nn =
        nn_function(vars_net_coordinates, homeless_coordinates,3)
      )

#plot multiple maps using for loop
vars_net.long.nn <- 
  dplyr::select(vars_net, ends_with(".nn")) %>%
    gather(Variable, value, -geometry)

vars <- unique(vars_net.long.nn$Variable)
mapList <- list()

for(i in vars){
  mapList[[i]] <- 
    ggplot() +
      geom_sf(data = filter(vars_net.long.nn, Variable == i), aes(fill=value), colour=NA) +
      scale_fill_viridis(name="") +
      labs(title=i) +
      theme_classic()}

do.call(grid.arrange,c(mapList, ncol = 3, top = "Nearest Neighbor risk Factors by Fishnet"))
```

